Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-06-28 manuscript.Rmd
Updates

Our tutorial is directed at researchers who have a basic \Revision{conceptual} understanding of Bayesian inference and parameter estimation and are planning to perform more complex Bayesian analyses \Revision{of psychological phenomena. We hope to achieve a balance of detail and overview---while we want to provide guidance for a full Bayesian workflow, similar to Gelman et al.} [-@gelman2020bayesian] \Revision{and Schad et al.}[-@schad2021toward; -@schad2021workflow]\Revision{, we don't shy away from in-depth explanations for the most curious scientists. To ensure that a broad audience is able to follow the tutorial, we limit and carefully explain equations. However, rudimentary knowledge of calculus is advantageous to fully understand Bayesian analysis. }

@@

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend [Rouder, Morey, & Pratte, 2013; Haaf & Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, -@van2021bayes]. This is shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for the interdependence between observations} [@van2021bayes; @van2023bayes]. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

\Revision{In hierarchical analyses, the general or overall effect is sometimes called fixed effect, and the individual-specific deviations from this effect are in that case referred to as random effects (Rouder \& Lu, 2005). We can also estimate individual effects, not as deviations from an overarching effect, but as effects for each individual, as obtained from a per-participant analysis. Throughout this tutorial, we will use the terms that suit the interpretation of our analyses, namely general effect, individual deviation, and individual effect.}

@@

\begin{align*}
&
\begin{aligned}
w_j = 
\begin{cases}
1 & j = 4\\
0 & \text{Otherwise},
\end{cases}
\end{aligned}
&
\begin{aligned}
z_j = 
\begin{cases}
1 & j = 3\\
0 & \text{Otherwise}.
\end{cases}
\end{aligned}
\end{align*}
\textcolor{blue}{The model equation shows that, for instance, for participant $i = 1$ and digit $j = 7$, with $\gamma_{i = 1}$ equal to $0.5$, $\beta_{i = 1}$ equal to $0.2$, $\delta_{7, i = 1}$ and $\delta_{3, i = 1}$ equal to $0.25$, and $\delta_{6, i = 1}$ and $\delta_{4, i = 1}$ equal to $0.3$, the mean RT is modeled by: $0.5 - 0.1 +  0.25$. If all $\delta$s are positive, then the model equation indicates that individuals respond fastest when the digits 2 and 8 are presented, as in this situation there are no digit effects in the equation. However, for digit $j = 8$, the mean RT is modeled by $0.5 - 0.1$ in this scenario, while for digit $j = 2$, the mean RT is modeled by $0.5 + 0.1$.}

@@

\begin{equation}
Y_{ijk} \sim \text{LogNormal}(\gamma_{i} + x_j \beta_{i} + u_j \delta_{7i} + v_j \delta_{6i} + w_j \delta_{4i} + z_j \delta_{3i}, \sigma^2).  
\end{equation}
Note that the only change in the model formulation is the change of the likelihood function from a normal distribution to a log-normal distribution. \Revision{The likelihood function describes the probability model that is assumed to be underlying the data. When the parameters of this probability model are known, the joint probability of the data can be inferred from the likelihood function. The log-normal distribution only assigns probability to positive data values, and is, just as the RT distributions, right-skewed. Therefore, the log-normal distribution is a better match for the RT data.} The location parameter of the log-normal distribution is described through the same combination of parameters as the mean of the normal model. However, it has a different interpretation. For example, for a location parameter of $\mu = 0$ and a scale parameter of $\sigma^2 = 1$, the mode of the log-normal distribution is $0.368$, and the mean is $1.65$.

@@

In summary, the models are placed directly on the raw RT data. The only difference between the normal and the log-normal models is the likelihood distribution. The normal model just assumes a symmetric normal distribution that can take on any value. The log-normal model assumes a log-normal distribution, which is restricted to positive values and has a right skew.

@@

Another important issue is that wide priors are often \Revision{improper probability distributions} [@hobert1996effect]. \Revision{A prior is proper if two conditions hold: 1. All values of the probability distribution are positive definite, and 2. the probability distribution sums to one for discrete data or integrates to one for continuous data. An example of a proper and improper probability distribution are provided in Figure} \@ref(fig:properfigure). \Revision{The figure shows two hypothetical priors for parameter $\theta$. The first condition can be assessed by checking the y-values---both priors have only positive definite values as function output (0.1, 0.2, and 0.4). To assess the second condition, we need to sum up the function values. Panel A shows a proper distribution where the values sum to 1; panel B shows an improper distribution where the values sum to 1.2.}

@@

The simulation proceeds as follows: First, one random draw from each prior distribution is taken. This is plugged into the individual effect distributions (Equation (2)). Then, the likelihood of the model, such as specified in Equations (1) and (3), is used to simulate observations per condition. This means that for *n* participants, *n* random values for the model parameters are generated based on the priors, which will result in the *k* observations per *j* conditions. This process is repeated *m* times, where *m* is the number of simulations drawn. Note that the more simulations are conducted, the closer one gets to the full picture because enough samples from the prior distribution are needed. Finally, we can visualize the simulated data. Figure \@ref(fig:priorpredfigurepanel) shows prior predictions for the key effects from the digit classification task. For this task, we mainly care about the side effect and digit effects corresponding to the $\beta$s and the $\delta$s from the model. Here, the prior predictions for the average side effect and the average digit effect across participants are depicted.

@@

In the symbolic distance model, we have to specify priors for the parameters $\mu_{\gamma}$, $\mu_{\beta}$, $\mu_{\delta}$, $\sigma^2_{\gamma}$, $\sigma^2_{\beta}$, and $\sigma^2_{\delta}$. The parameters $\mu_{\gamma}$, $\mu_{\beta}$, and $\mu_{\delta}$ can be positive or negative. All variance terms ($\sigma_{\gamma}^2$, $\sigma_{\beta}^2$, $\sigma_{\delta}^2$, and $\sigma^2$) are constrained to be positive. If there is no information from previous studies, researchers might be inclined to use the default priors in `brms` as a starting point.

In `brms`, different default priors are set on different types of parameters [@R-brms_a]. The prior on the group-level regression weights ($\mu_{\beta}$, $\mu_{\delta}$) is a normal distribution with a mean of 0 and a standard deviation of 1. The prior on the intercept $\mu_{\gamma}$ is a scaled and shifted Student's *t*-distribution with 3 degrees of freedom, a location parameter of 1, and a scale parameter of 10. `brms` specifies priors on standard deviations ($\sqrt{\sigma^2}$) instead of variances. For the normal model, this means that priors are set on $\sigma_{\gamma}$, $\sigma_{\beta}$, $\sigma_{\delta}$ and $\sigma$, instead of on $\sigma_{\gamma}^2$, $\sigma_{\beta}^2$, $\sigma_{\delta}^2$ and $\sigma^2$. The default prior on the standard deviations is a central, scaled Student's *t*-distribution with 3 degrees of freedom and a scale parameter of 10 that is truncated at zero. To summarize, the default priors would be set the following way:

\begin{equation}
\begin{aligned}
\mu_{\gamma} &\sim \text{Student's t}(3, 1, 10), (\#eq:chosenpriors1) \\ 
\mu_{\beta}, \mu_{\delta} &\sim \text{Normal}(0, 1), \\ 
\sigma, \sigma_{\gamma}, \sigma_{\beta}, \sigma_{\delta} &\sim \text{Truncated-Student's t}_{+}(3, 0, 10). 
\end{aligned}
\end{equation}
The prior distributions `brms` specifies on the standard deviations can be transformed to prior distributions on variances ($var$). Prior distributions on variances have a smaller scale than priors on standard deviation ($sd$). Note, however, that the transformation of the prior distributions is not as simple as the transformation of the parameter itself. \Revision{If we formulate a prior on standard deviations, there is always an implied prior on the variances as well. If we think of the distribution as $f(x)$, $x$ can either be defined as $\sqrt{(var)}$, if we formulated the prior on the standard deviation, or $x$ can be defined as ${sd}^2$ if we formulated the prior on the variance. From this, we can derive the implied distribution $f(var)$ or $f(sd)$. Notably, this is not the same as calculating $\sqrt{(f(x))}$ or $f(x)^2$. To transform the priors on standard deviations to variances, we try to find the parameters that match the prior distribution to the implied distribution.} In [Online Supplement J](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/R%20code/Prior%20transformation), we provide R-code with a numeric solution to the transformation. For the `brms` priors, we obtain the following implied prior distributions on variances:

@@

The figure shows the difference in RTs between digits for every repetition averaged across subjects (i.e., 1000 grey dots per effect). The blue dots represent the mean effect. Using the default prior from `brms`, the mean effects are expected to be around zero, with a mean effect of zero. However, the range around this mean is very wide, ranging from -10 to 10 second effects. Most of the trials are in a smaller range of -5 to 5 second effects. This means that according to our priors, we would expect the effects to fall in this range. For this task, where participants have to respond as fast as possible, such large RT effects are unlikely. Therefore, it seems reasonable to adjust the priors.

@@

The normal distribution on $\mu_{\beta}$ and $\mu_{\delta}$ seems reasonable if we assume that a two-sided hypothesis about the symbolic distance effect is tested. However, we expect less variance around this mean. Therefore, we decrease the setting for the variation. The wide tails of the Student's *t*-distribution could be one of the reasons for the high range in variation. We change the prior for $\mu_{\gamma}$ to a normal distribution that has more narrow tails. As RTs cannot be negative, we adjust the prior to be positive only by truncating the normal distribution using a lower bound of zero. Finally, we would like to use a more common conjugate prior for the variances to ensure obtaining a proper posterior distribution. ^[Conjugate priors result in posterior distributions that belong to the same distribution family as the prior distribution [@Fink1997]. For a normal likelihood, these are the normal distribution for the mean and the inverse gamma distribution for the variance.] ^[\Revision{However, in rstan, non-conjugate priors are commonly applied} [e.g., @gelman2006prior].] Therefore, we choose an inverse gamma distribution. We can also be more specific about the expected RTs. Based on our knowledge of these types of tasks [e.g., @haaf2017developing], it seems highly plausible that participants will respond within a second or two. We would expect effects in the range of 10 milliseconds to 100 milliseconds. We adjust our priors using these new distribution types and settings, evaluate the prior predictions (shown in panel B of Figure \@ref(fig:priorpredfigurepanel)) and arrive at the following priors:^[Because response times cannot be negative the prior for the overall baseline $\mu_\gamma$ is a truncated normal with a lower bound of zero.]

@@

```{r priorpredfigurepanellog, echo=FALSE, fig.cap="Predicted mean RT difference between digits per repetition. The blue line represents the aggregated trend. Panel A represents the prior prediction based on the priors of the normal model. Panel B shows the prior prediction based on the selected prior for the log-normal model.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="100%", eval = FALSE}
knitr::include_graphics("I - Images/priorpredfigure_lognormal_v2.png")
```
The transformation of the prior distributions is not as simple as the transformation of the parameter itself. Panel C shows that with the same priors as for the normal model, we obtain different data for the log-normal model compared to the normal model (panel B). Therefore, we adjusted the priors until the simulated data matched our expectations and looked like panel B. This yields the following priors that result in the prior prediction shown in panel D of Figure \@ref(fig:priorpredfigurepanel):^[For the overall baseline $\mu_\gamma$ we adjusted the prior based on our expectation of the expected overall response times. A mean parameter of -0.5 for the log-normal likelihood corresponds to a mean response time of 1.]

@@

Every sampling algorithm requires the specification of the number of samples (also called iterations) that are drawn from the posterior distribution. Furthermore, all MCMC sampling methods have in common that results depend (somewhat) on chosen starting values, that is, initial parameter values for which the unnormalized posterior (i.e., likelihood times prior) is evaluated. This dependency can be assessed if the sampling algorithm is repeated several times (i.e., chains) with different starting values. Since the first few iterations of a chain---often called warm-up or burn-in---are somewhat dependent on the starting value, these samples are typically discarded.

@@

`Rstan` allows the application of `Stan` [@carpenter2017stan], a probabilistic programming language, in `R`. The package uses a version of the Hamiltonian Monte Carlo (HMC) algorithm, No-U-turn sampler [NUTS; Hoffman & Gelman, -@hoffman2014no], for sampling. `Rstan` requires two files: a `.stan` file containing the model specification and an R script containing the code for the model fitting. In the `.stan` file, the parameters, priors, and the likelihood are specified. An example `.stan` file of a simple, non-hierarchical normal model, where the variance is already known and the mean has to be estimated, is shown below [adapted example from @nicenboim2021introduction, Chapter 10]. The `.stan` file for the symbolic distance effect, including explanations, can be found in [Online Supplement K](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/R%20objects/rstan/normal%20model/myModel.stan). The file is divided into three sections: `data`, `parameters`, and `model`.

@@

model {
  // Priors: 
  target += normal_lpdf(mu | 0, 20);
  // Likelihood: 
  for(i in 1:N)
    target += normal_lpdf(y[i] | mu, sigma);
}
```
\normalsize

The `data` section specifies the variables in the dataset, as well as constant values in the model. For the symbolic distance effect model, the `data` section contains the total number of observations, the participant index variable, the response time variable in the dataset, the condition for every response time (side of the digit and the digit indicator, i.e., variables $x_j$, $u_j$, $v_j$, $w_j$ and $z_j$ from \@ref(eq:datalevel)) and the prior specification according to Equations \@ref(eq:chosenpriors2var) and \@ref(eq:chosenpriorslogvar). The `parameters` section specifies all parameters that are estimated. These are the general and individual effects. The last section, `model`, contains the likelihood function and priors, as specified earlier in this article.

@@

The package `rstan` offers a lot of flexibility. It is possible to specify the likelihood and priors in the exact way as is wished. However, this comes with the price of some technical knowledge. The commented code for fitting the full symbolic distance effect model using `rstan`, can be found in [Online Supplement C: *Bayesian Hierarchical Modeling in Rstan*](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/C%20-%20Tutorial%20Normal%20Model%20Rstan/Online-Supplement-C---Bayesian-Hierarchical-Modeling-in-rstan.pdf).

@@

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \@ref(fig:traceplots3p). For the package `rstan` the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The `brms` package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the `rstan` model. If the number of effective samples is too low, `brms` and `rstan` will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \Revision{The difference between the model diagnostics of} `rstan` \Revision{and} `brms` \Revision{can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). Although we tried to make the models as similar as possible, they are not identical.}

@@

Imagine this results in $\text{BF}_{\text{nl}} = 100$. This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \Revision{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 fulfills?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. [-@rouder2018theories] argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

@@

A possible explanation is a lack of individual variability in the digit and side effects. The full model assumes that participants vary in the digit and side effect, and we specified our expectations about this size of the variability in the prior. However, if there is a lack of individual variability in the effects, the full model is punished for the complexity added by the individual variation (illustrated by Rouder et al., 2021). To investigate this idea, we can reformulate the prior distributions on the variance parameters to indicate a smaller individual variation or, even more extreme, remove these individual effects from the model entirely. Then, we can calculate the posterior model probabilities for all specified models.

>>>>

Our tutorial is directed at researchers who have a basic \Revision{conceptual} understanding of Bayesian inference and parameter estimation and are planning to perform more complex Bayesian analyses \Revision{of psychological phenomena. We hope to achieve a balance of detail and overview---while we want to provide guidance for a full Bayesian workflow, similar to Gelman et al.} [-@gelman2020bayesian] \Revision{and Schad et al.}[-@schad2021toward; -@schad2021workflow]\Revision{, we don't shy away from in-depth explanations for the most curious scientists. To ensure that a broad audience is able to follow the tutorial, we limit and carefully explain equations. However, rudimentary knowledge of calculus is advantageous to fully understand the Bayesian analyses. }

@@

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend [Rouder, Morey, & Pratte, 2013; Haaf & Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, -@van2021bayes].

\Revision{A pattern where a general trend is not representing the individual trend is} shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for differential experiment effects} [@van2021bayes; @van2023bayes]. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

\Revision{In hierarchical analyses, the general or overall effect is sometimes called fixed effect, and the individual-specific deviations from this effect are in that case referred to as random effects (Rouder \& Lu, 2005). We can also estimate individual effects, not as deviations from an overarching effect, but as effects for each individual, as obtained from a per-participant analysis. Throughout this tutorial, we will use the terms general effect, individual deviation, and individual effect.}

@@

\begin{align*}
&
\begin{aligned}
w_j = 
\begin{cases}
1 & j = 4\\
0 & \text{Otherwise},
\end{cases}
\end{aligned}
&
\begin{aligned}
z_j = 
\begin{cases}
1 & j = 3\\
0 & \text{Otherwise}.
\end{cases}
\end{aligned}
\end{align*}

\Revision{The model equation enables us to compute the expected response time $Y_{ij}$ for a participant $i$ when presented with a specific probe. For example, if participant $i = 1$ is presented with digit $j = 7$, $x_j$ takes the value $-\frac{1}{2}$ and $u_j$ takes the value $1$. The parameters $v_j$, $w_j$, and $z_j$ take the value zero, such that the equation for the expected value reduces to: $\gamma_{i} + \frac{1}{2} \beta_{i} + \delta_{7,i}$. Let us assume that the participant 1's mean response time across all items is $\gamma_{1} = 0.5$, they respond generally faster to digits to the right of five, as expressed by $\beta_{1} = 0.2$, and they respond slower to digit 7, given by $\delta_{7,1} = 0.25$. Then their expected response time for the digit 7 can be calculated by $0.5 - 0.5 * 0.2 + 1 * 0.25$.}

@@

\begin{equation}
Y_{ijk} \sim \text{LogNormal}(\gamma_{i} + x_j \beta_{i} + u_j \delta_{7i} + v_j \delta_{6i} + w_j \delta_{4i} + z_j \delta_{3i}, \sigma^2).  
\end{equation}
Note that the only change in the model formulation is the change from a normal distribution to a log-normal distribution. \Revision{The density function describes the probability model that is assumed to be underlying the data. When the parameters of this probability model are known, the joint probability of the data can be inferred from the \Revision{density} function. The log-normal distribution only assigns probability to positive data values, and is, just as the RT distributions, right-skewed. Therefore, the log-normal distribution is a better match for the RT data.} The location parameter of the log-normal distribution is described through the same combination of parameters as the mean of the normal model. However, it has a different interpretation. For example, for a location parameter of $\mu = 0$ and a scale parameter of $\sigma^2 = 1$, the mode of the log-normal distribution is $0.368$, and the mean is $1.65$.

@@

In summary, the models are placed directly on the raw RT data. The only difference between the normal and the log-normal models is the probability distribution. The normal model just assumes a symmetric normal distribution that can take on any value. The log-normal model assumes a log-normal distribution, which is restricted to positive values and has a right skew.

@@

Another important issue is that wide priors are often \Revision{improper probability distributions} [@hobert1996effect]. \Revision{A prior is proper if two conditions hold: 1. All values of the probability distribution are equal or greater than zero, and 2. the probability distribution sums to one for discrete data or integrates to one for continuous data. An example of a proper and improper probability distribution are provided in Figure} \@ref(fig:properfigure). \Revision{The figure shows two hypothetical priors for parameter $\theta$. The first condition can be assessed by checking the y-values---both priors have only positive values as function output (0.1, 0.2, and 0.4). To assess the second condition, we need to sum up the function values. Panel A shows a proper distribution where the values sum to 1; panel B shows an improper distribution where the values sum to 1.2.}

@@

The simulation proceeds as follows: First, one random draw from each prior distribution is taken. This is plugged into the individual effect distributions (Equation (2)). Then, the probability of the model, such as specified in Equations (1) and (3), is used to simulate observations per condition. This means that for *n* participants, *n* random values for the model parameters are generated based on the priors, which will result in the *k* observations per *j* conditions. This process is repeated *m* times, where *m* is the number of simulations drawn. Note that the more simulations are conducted, the closer one gets to the full picture because enough samples from the prior distribution are needed. Finally, we can visualize the simulated data. Figure \@ref(fig:priorpredfigurepanel) shows prior predictions for the key effects from the digit classification task. For this task, we mainly care about the side effect and digit effects corresponding to the $\beta$s and the $\delta$s from the model. Here, the prior predictions for the average side effect and the average digit effect across participants are depicted.

@@

In the symbolic distance model, we have to specify priors for the parameters $\mu_{\gamma}$, $\mu_{\beta}$, $\mu_{\delta}$, $\sigma^2_{\gamma}$, $\sigma^2_{\beta}$, and $\sigma^2_{\delta}$. The parameters $\mu_{\gamma}$, $\mu_{\beta}$, and $\mu_{\delta}$ can be positive or negative. All variance terms ($\sigma_{\gamma}^2$, $\sigma_{\beta}^2$, $\sigma_{\delta}^2$, and $\sigma^2$) are constrained to be positive. If there is no information from previous studies, researchers might be inclined to use the default priors \Revision{of software packages, often reasonably uninformative priors,} as a starting point.

\Revision{We take the reasonably uninformative priors as a starting point as well specifying priors on different types of parameters. These priors were previously used as} default priors in `brms` [@R-brms_a]. The prior on the group-level regression weights ($\mu_{\beta}$, $\mu_{\delta}$) is a normal distribution with a mean of 0 and a standard deviation of 1. The prior on the intercept $\mu_{\gamma}$ is a scaled and shifted Student's *t*-distribution with 3 degrees of freedom, a location parameter of 1, and a scale parameter of 10. `brms` specifies priors on standard deviations ($\sqrt{\sigma^2}$) instead of variances. For the normal model, this means that priors are set on $\sigma_{\gamma}$, $\sigma_{\beta}$, $\sigma_{\delta}$ and $\sigma$, instead of on $\sigma_{\gamma}^2$, $\sigma_{\beta}^2$, $\sigma_{\delta}^2$ and $\sigma^2$. The default prior on the standard deviations is a central, scaled Student's *t*-distribution with 3 degrees of freedom and a scale parameter of 10 that is truncated at zero. To summarize, the default priors would be set the following way:

\begin{equation}
\begin{aligned}
\mu_{\gamma} &\sim \text{Student's t}(3, 1, 10), (\#eq:chosenpriors1) \\ 
\mu_{\beta}, \mu_{\delta} &\sim \text{Normal}(0, 1), \\ 
\sigma, \sigma_{\gamma}, \sigma_{\beta}, \sigma_{\delta} &\sim \text{Truncated-Student's t}_{+}(3, 0, 10). 
\end{aligned}
\end{equation}
The prior distributions specified on the standard deviations can be transformed to prior distributions on variances ($var$). Prior distributions on variances have a smaller scale than priors on standard deviation ($sd$). Note, however, that the transformation of the prior distributions is not as simple as the transformation of the parameter itself. \Revision{If we formulate a prior on standard deviations, there is always an implied prior on the variances as well. If we think of the distribution as $f(x)$, $x$ can either be defined as $\sqrt{(var)}$, if we formulated the prior on the standard deviation, or $x$ can be defined as ${sd}^2$ if we formulated the prior on the variance. From this, we can derive the implied distribution $f(var)$ or $f(sd)$. Notably, this is not the same as calculating $\sqrt{(f(x))}$ or $f(x)^2$. To transform the priors on standard deviations to variances, we try to find the parameters that match the prior distribution to the implied distribution.} In [Online Supplement J](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/R%20code/Prior%20transformation), we provide R-code with a numeric solution to the transformation. For the `brms` priors, we obtain the following implied prior distributions on variances:

@@

The figure shows the difference in RTs between digits for every repetition averaged across subjects (i.e., 1000 grey dots per effect). The blue dots represent the mean effect. Using the \Revision{reasonably uninformatieve priors}, the mean effects are expected to be around zero, with a mean effect of zero. However, the range around this mean is very wide, ranging from -10 to 10 second effects. Most of the trials are in a smaller range of -5 to 5 second effects. This means that according to our priors, we would expect the effects to fall in this range. For this task, where participants have to respond as fast as possible, such large RT effects are unlikely. Therefore, it seems reasonable to adjust the priors.

@@

The normal distribution on $\mu_{\beta}$ and $\mu_{\delta}$ seems reasonable if we assume that a two-sided hypothesis about the symbolic distance effect is tested. However, we expect less variance around this mean. Therefore, we decrease the setting for the variation. The wide tails of the Student's *t*-distribution could be one of the reasons for the high range in variation. We change the prior for $\mu_{\gamma}$ to a normal distribution that has more narrow tails. As RTs cannot be negative, we adjust the prior to be positive only by truncating the normal distribution using a lower bound of zero. Finally, we would like to use a more common conjugate prior for the variances to ensure obtaining a proper posterior distribution. ^[Conjugate priors result in posterior distributions that belong to the same distribution family as the prior distribution [@Fink1997]. For a normal model, these are the normal distribution for the mean and the inverse gamma distribution for the variance.] ^[\Revision{However, in rstan, non-conjugate priors are commonly applied} [e.g., @gelman2006prior].] Therefore, we choose an inverse gamma distribution. We can also be more specific about the expected RTs. Based on our knowledge of these types of tasks [e.g., @haaf2017developing], it seems highly plausible that participants will respond within a second or two. We would expect effects in the range of 10 milliseconds to 100 milliseconds. We adjust our priors using these new distribution types and settings, evaluate the prior predictions (shown in panel B of Figure \@ref(fig:priorpredfigurepanel)) and arrive at the following priors:^[Because response times cannot be negative the prior for the overall baseline $\mu_\gamma$ is a truncated normal with a lower bound of zero.]

@@

```{r priorpredfigurepanellog, echo=FALSE, fig.cap="Predicted mean RT difference between digits per repetition. The blue line represents the aggregated trend. Panel A represents the prior prediction based on the priors of the normal model. Panel B shows the prior prediction based on the selected prior for the log-normal model.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="100%", eval = FALSE}
knitr::include_graphics("I - Images/priorpredfigure_lognormal_v2.png")
```
The transformation of the prior distributions is not as simple as the transformation of the parameter itself. Panel C shows that with the same priors as for the normal model, we obtain different data for the log-normal model compared to the normal model (panel B). Therefore, we adjusted the priors until the simulated data matched our expectations and looked like panel B. This yields the following priors that result in the prior prediction shown in panel D of Figure \@ref(fig:priorpredfigurepanel):^[For the overall baseline $\mu_\gamma$ we adjusted the prior based on our expectation of the expected overall response times. A mean parameter of -0.5 for the log-normal distribution corresponds to a mean response time of 1.]

@@

Every sampling algorithm requires the specification of the number of samples (also called iterations) that are drawn from the posterior distribution. Furthermore, all MCMC sampling methods have in common that results depend (somewhat) on chosen starting values, that is, initial parameter values that \Revision{are fed to the sampler in the first iteration}. This dependency can be assessed if the sampling algorithm is repeated several times (i.e., chains) with different starting values. Since the first few iterations of a chain---often called warm-up or burn-in---are somewhat dependent on the starting value, these samples are typically discarded.

@@

`Rstan` allows the application of `Stan` [@carpenter2017stan], a probabilistic programming language, in `R`. The package uses a version of the Hamiltonian Monte Carlo (HMC) algorithm, No-U-turn sampler [NUTS; Hoffman & Gelman, -@hoffman2014no], for sampling. `Rstan` requires two files: a `.stan` file containing the model specification and an R script containing the code for the model fitting. In the `.stan` file, the parameters, priors, and the \Revision{probability distribution of the data} are specified. An example `.stan` file of a simple, non-hierarchical normal model, where the variance is already known and the mean has to be estimated, is shown below [adapted example from @nicenboim2021introduction, Chapter 10]. The file is divided into three sections: `data`, `parameters`, and `model`.

@@

model {
  // Priors: 
  target += normal_lpdf(mu | 0, 20);
  // Probability model: 
  for(i in 1:N)
    target += normal_lpdf(y[i] | mu, sigma);
}
```
\normalsize

The `data` section specifies the variables in the dataset, as well as constant values in the model. For the symbolic distance effect model, the `data` section contains the total number of observations, the participant index variable, the response time variable in the dataset, the condition for every response time (side of the digit and the digit indicator, i.e., variables $x_j$, $u_j$, $v_j$, $w_j$ and $z_j$ from \@ref(eq:datalevel)) and the prior specification according to Equations \@ref(eq:chosenpriors2var) and \@ref(eq:chosenpriorslogvar). The `parameters` section specifies all parameters that are estimated. These are the general and individual effects. The last section, `model`, contains the probability function and priors, as specified earlier in this article. \Revision{Instead of using} `y[i] ~ normal(mu, sigma)` \Revision{to estimate parameters, such as the mean from the example, we used the} `target` \Revision{specification} [for a in-depth explanation see @nicenboim2021introduction, Chapter 10]. \Revision{Both methods indicate in this case that} `y[i]` \Revision{is normally distributed with the mean} `mu` \Revision{and the variance} `sigma`. \Revision{However, the target method saves constant terms that are required for model comparison} [@R-bridgesampling] The `.stan` file for the symbolic distance effect, including explanations, can be found in [Online Supplement K](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/R%20objects/rstan/normal%20model/myModel.stan).

@@

The package `rstan` offers a lot of flexibility. It is possible to specify the probability model and priors in the exact way as is wished. However, this comes with the price of some technical knowledge. The commented code for fitting the full symbolic distance effect model using `rstan`, can be found in [Online Supplement C: *Bayesian Hierarchical Modeling in Rstan*](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/C%20-%20Tutorial%20Normal%20Model%20Rstan/Online-Supplement-C---Bayesian-Hierarchical-Modeling-in-rstan.pdf).

@@

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \@ref(fig:traceplots3p). For the package `rstan` the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The `brms` package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the `rstan` model. If the number of effective samples is too low, `brms` and `rstan` will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \Revision{The difference between the model diagnostics of} `rstan` \Revision{and} `brms` \Revision{can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). We equated the models where sensible which resulted in models that are fairly comparable. }

@@

Imagine this results in $\text{BF}_{\text{nl}} = 100$. This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \Revision{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 is big enough?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. [-@rouder2018theories] argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

@@

A possible explanation is a lack of individual variability in the digit and side effects. The full model assumes that participants vary in the digit and side effect, and we specified our expectations about this size of the variability in the prior. However, if there is a lack of individual variability in the effects, the full model is punished for the complexity added by the individual variation (illustrated by Rouder et al., 2021). To investigate this idea, we can reformulate the prior distributions on the variance parameters to indicate a smaller individual variation or, even more extreme, remove these individual effects from the model entirely. Then, we can calculate the posterior model probabilities for all specified models. \Revision{Typically in sensitivity analyses, the sensitivity of the Bayes factor on the width of the prior on effect size is assessed} [e.g., @van2019jasp]\Revision{. However, in our case the prior distributions on the variance parameters are key to understand sensitivity.}
